Combinatorial Mixture Models for Single Cell Assays with Application to Vaccine Studies

### 2012-07-20 MIMOSA.tex

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065\footnote{Why do we change data now?} data set (Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, MIMOSA outperformed other approaches.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.3\columnwidth]{Figures/simulations_onesided_AUCS.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_AUC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (0,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (4.5,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_onesided_FDR.pdf}};
   \node[anchor=south west, inner sep=0] at (9,-3.5){\includegraphics[width=0.3\columnwidth]{Figures/simulations_twosided_FDR.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
    \node at (0,-0.25) [font=\tiny\sffamily] {D} ;
    \node at (4.6,-0.25) [font=\tiny\sffamily] {E} ;
    \node at (9,-0.25) [font=\tiny\sffamily] {F} ;
\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one--sided model (top row) and the two--sided model (bottom row). Ten simulations were generated at each of seven increasing values of $N$ (total counts) using hyper--parameter estimates from real ICS data. A) Boxplots of the area under the ROC curves for Fisher's exact test, MIMOSA (MCMC constrained model), likelihood ratio test, and the ranked log fold change. B) Boxplots for the unconstrained model. C) Observed and nominal false discovery rate for a representative simulated data set (N=100,000 counts) MIMOSA (one--sided), Fisher's exact test, and the likelihood ratio test. C) ROC curves for a representative simulated data set (N=100,000 counts). D) AUC  boxplots, E) observed vs. nominal FDR, F) ROC analysis comparing MIMOSA (MCMC unconstrained model) and competing methods for two--sided simulated data. }\label{fig:simulations}
\end{figure}

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_AUC.pdf}};
    \node[anchor=south west, inner sep=0] at (4.5,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_ROC.pdf}};
   \node[anchor=south west, inner sep=0] at (9,0) {\includegraphics[width=0.3\columnwidth]{Figures/Sim_Trunc_FDR.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
    \node at (9,3.4) [font=\tiny\sffamily] {C} ;
 \end{tikzpicture}
   \caption{Unconstrained MIMOSA model fit to data where model assumptions are violated. Two--sided data were simulated from a model where the proportions were drawn from a truncated normal distribution over $[0,1]$, rather than a Beta distribution.}
   \label{fig:simulations_trunc}
\end{figure}

>>>>

To assess the sensitivity of the model to deviations from model assumptions, we repeated the simulations with the cell proportions drawn from  truncated normal distributions on $(0,1)$, rather than beta distributions. The means and variances of the truncated normal distributions were set to the maximum likelihood estimates of the beta distributions defined by the $\alpha,\beta$ hyper parameters estimated from the HVTN065 data set (Supplementary Figure~\ref{fig:simulations_trunc}). Even under these departures from the model assumptions, the unconstrained MIMOSA model outperformed Fisher's exact test and performed about as well as the constrained MIMOSA model fit to one-sided data.

@@

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
\begin{tikzpicture}
    \node[anchor=south west,inner sep=0] at (0,0) {\includegraphics[width=.4\columnwidth]{Figures/Sim_onesided_ROC_5000.pdf}};
    \node[anchor=south west, inner sep=0] at (6,0){\includegraphics[width=0.4\columnwidth]{Figures/Sim_onesided_FDR_5000.pdf}};
    \node at (0,3.4) [font=\tiny\sffamily] {A} ;
    \node at (4.6,3.4) [font=\tiny\sffamily] {B} ;
\end{tikzpicture}
   \caption{Comparison of positivity detection methods on data simulated from the one--sided model. Ten simulations were generated at an $N$ of 5,000 total counts using hyper--parameter estimates from real ICS data (IFNg expressing CD4+ T--cells stimulated with ENV-1-PTEG from HVTN065) with a five--fold effect size between responder and non--responder components. A) Average ROC curve over the 10 simulated data sets (N=5,000) D) Average observed and nominal false discovery rate over 10 simulated data sets (N=5,000). Curves are shown for MIMOSA (one--sided), Fisher's exact test, the likelihood ratio test, and ranked log fold--change. Results for the two--sided model and the MIMOSA fit to a model violating model assumptions, as well as other values of N are in the Supplementary Material.}\label{fig:simulations}
\end{figure}
